Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2020-06-12 main.Rmd
add discussion

We have compiled a set of known experimentally validated effectors from phytopathogenic bacteria, fungi and oomycetes and used them as positive training examples with which to tune a range of CNN classifiers.

>>>>

We have compiled a set of known experimentally validated effectors from phytopathogenic bacteria, fungi and oomycetes and used them as positive training examples with which to tune a range of CNN classifiers, the sequences are all taken from the database PHI-Base, a manually curated database that aims to search all the literature on pathogen and host interactions [@phibase]. The data in PHI-Base is complete as far as 2013 at the moment, thus the phytopathogen effectors we have collected should be all those shown experimentally to have an effect on a host that have been reported on since that time. We chose this set as we believed that this would give us the most reliable and unbiased set of effectors on which we could train learners. That is not to say that the set itself cannot be biased and that the set does not introduce any bias into the classifications of our learners. Sources of bias in our sequence set include the time limits on what has so far been included in PHI-Base, any effectors known but not reported on in this seven year time period cannot be represented in the models. The species of phytopathogens represented in the set also create bias, the effectors are not selected from species sampled proportionally or randomly but instead are those that trends in effector research over the last seven years have brought to focus. In particular, species that have had genome sequencing projects over this time are over-represented. There may also be some echoes of other methods previously applied, effectors studied experimentally must be identified first as hypothetical effectors, usually with the aid of computational tools whose models are themselves biased towards the sort of sequence that we already know. The effectors in the literature may therefore be enriched with respect to known features and results from classifiers should be interpreted with this in mind. We are not easily able to quantify any bias, but the greater generalisability of our models over the others tested gives us reason to believe that the sequence set trained on was broad enough for good models to be developed.

A common misconception of deep learning models is that training datasets need to be extremely large, this is only half true. In fact, in practice training data need only to be large with respect to the model size. Here we have coupled small data sets (tens or hundreds of training examples) with small models. The layers and architectures of the models presented here are much smaller than those typically used in large machine vision projects for example. Yet the small models remain useful and have predictive power, indicating a definite role for deep learning approaches in the work of bioinformaticians without truly massive datasets.

Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised and although only a small fraction ^[Ruth, have we any idea what this fraction might be] of the possible hyperparameter and parameter space was explored we still compared 6000 models. At a run time of around ^[Ruth, approx run time per model] So even for relatively small data sets, a large amount of specialised GPU compute power is required.

The models we created performed exceptionally on the PHI-Base hold-out validation data set of phytopathogenic effector sequences. The greater than 80 % classification accuracy for the fungal and oomycete models is an excellent accuracy on such sequences and our models outperformed the other classifiers tested by large margins on the PHI-Base fungal, no other machine learning method has been reported to have performed as well on phytopathogen sequences. The greater than 97 % accuracy we observed in our model trained on bacterial phytopathogen effectors is also exceptional and similar to what Xue et al [@deept3] showed in human bacterial pathogen effectors. When we evaluated the proportion of effectors each of our and other classifiers could find in data used to create the other classifiers, we found that our bacterial model outperformed the bacterial models again. A slightly different picture emerged when we compared our method with EffectorP 1 and 2 on fungal data. Both versions of Effector P outperformed our model on training data used to create Effector P in the first instance but identical predictions were made by all fungal classifiers on the validation set provided by Effector P 2. This was coincident with a large drop in accuracy from training to validation data by Effector P 1 and 2.  Combined with observations on our PHI-Base data on higher numbers of false positives from Effector P, we conclude that Effector P is over-fitted slightly on its training data and that our model is likely to be more generally accurate. We expect that the tools used in concert will provide very good classification and prediction of fungal effectors. Our model also performed worse on the data used to test the untrained heuristic oomycete RXLR sequence detector EffectR, finding three of six, relative to EffectR's five of six. This is likely due to the six test sequences being exclusively RXLR and our model being trained on effectors in general, the presence of RXLR not being diagnostic of effector-ness in our model means Researchers hoping to find exclusively RXLR containing proteins would be advised to use EffectR, those hoping to find effectors beyond RXLRs may find utility with our oomycete model.

In developing a negative training data set that contained secreted non-effector proteins we hoped to decouple the identification of effector from identification of secreted protein as a proxy for effector. We believe that we have been succesful at this, the models we developed do identify effectors against a background of secreted proteins, indicating that they have some internal representation of what makes an effector different from a secreted peptide. By examining the activation maps of the CNN models, we learned that the maximum activations in the models remains in the N-terminus of the proteins, coincident with the expected positions of secretions signals and is relatively narrow (about 25 amino acids. We also noted that there is no typical primary sequence motif that can be identified, the identity of the amino acids themselves does not seem important. We did find that various categories of amino acid were enriched or depleted significantly in the narrow activation region of effectors, relative to the same area of non-effector secreted proteins. A favouring of functional properties of the amino acids of the effector over the actual identity may be reflective of the many functional types origins of effectors. It may be that the N-terminal activation region in the effectors represents non-canonical secretion-compatible signal in the effector proteins. In order to evolve towards a useful secretion signal from an arbitrary point it may be enough to have a region that satisfies some chemical property that makes it secretable, yet is different enough in effectors to separate them from the secreted proteins by a deep learning algorithm.

Finally, we have made our models available in an R package _deepredeff_ - from 'deep learning prediction of effector'. This can be obtained from GitHub as shown in methods, installation and usage instructions are available in the documentation provided there. The R package allows the user to run the sequences from various sequence format files against a selected model and obtain a _p_-value that each sequence is an effector according to the model. Various summaries and plots are provided for the user when working on large numbers of sequences. The package integrates well with Bioconductor. For those wishing to use Python to make use of our models, we provide the models as HDF5 files from TensorFlow that can be use in that toolkit.
